Handling String-Number Conversion in String Constraint Solving

### 2019-11-12 draft.tex
polish under-approx

\begin{lemma}[\cite{SeidlSMH04}]
$\parikhof{A}$ of an FA $A$ can be computed in the form of a linear formula  $\parikhfof{A}$ such that $\modelsof{\parikhfof A} =_\Sigma \parikhof{A}$, in time linear to the size of $A$.
\end{lemma}

@@

%Such automata are recognizers of symbolic strings, i.e., strings over $\cvar$.
Parametric automata accept strings over $\cvar$, called \emph{parametric strings}, but we still use them as representations of languages over $\Sigma$. 
%
Namely, string over $\cvar$ are interpreted as strings over $\Sigma$ given an assignment of characters to its character variables.  
%
Formally, an \emph{interpretation $I$ over $\cvar$} is a mapping from $\cvar$ to $\Sigma_\epsilon$ and for a parametric string $x= v_1v_2\cdots v_k$ over $\cvar$, $I(x)$ is defined as $I(v_1)\cdot I(v_2)\cdot \ldots \cdot I(v_k)$.
%
We define the semantics of $x$ as the set $\semof x$ of all possible interpretations of $x$, the semantics of a language $L\subseteq V_\Sigma$ is the union $\semof L = \cup_{x\in L} \semof{x}$ of semantics of its words, and the semantics $\semof A$ of an PFA is the semantics of its language.

The crucial feature of parametric flat automata is that their semantics can be faithfully represented by a linear integer arithmetic formulas and handled efficiently by an SMT solver. 
\lh{, allowing to circumvent more costly automata algorithms and decision procedures. Concretize and justify this better?}
%
Achieving this was the purpose of introducing the structural constraints of PFA. 
Their implication central for our method is that every parametric string $x\in L(A)$ is uniquelly determined by its Parikh image $\parikhof{x}$, as formalised by the following lemma.

@@

Since every string $w \in \semof{A}$ is obtained from a parametric string $x\in L(A)$ and an interpretation of character varriables $I:\cvar\rightarrow\Sigma_\epsilon$ such that $w = I(x)$, and since $x = \decode A(\parikhof x)$, 
we may say that the assignment $I \cup \parikhof x$ uniting the interpretation $I$ and the Parikh image of the parametric string $x$ encodes $w$. Lemma~\ref{lemma:parikh} also implies that two different strings are encoded by different united assignments (the Parikh image or the interpretation must be different), and hence we may extend the decoding function $\decode A$ to united assignments so that $\decode A(I \cup x) = w$.
Encodings of strings in $\semof{A}$ are encodings of parametric strings of $L(A)$ paired with arbitrary interpretations of character variables $\cvar$, which are precisely all modesl of the formula $\parikhfof A$ interpreted over $\cvar\cup\pvarsof \cvar$, as formulated byt the following corollary of Lemma~\ref{lemma:decx}:
%
\begin{corollary}
For an PFA $A$ over character variables $V$,
$\semof{A} = \{\decode A(I\cup \pim) \mid I\cup\pim\in\semof{\parikhfof A}_{\cvar\cup\pvar}\}$.
\end{corollary}
%
A minor technical difficulty is that we wish to work with pure arithmetic formulas, 
but we may also wish to specify when a character variable is assigned $\epsilon$, which is not a number. 
Hence, to allow purely arithmetic formulas to talk about the value $\epsilon$, we encode $\epsilon$ as $-1$.
In the further development, $\epsilon$ will mean $-1$ when working with arithmetic formulas constraining interpretations of character variables.

\paragraph{Synchronisation of PFA}
The following construction is an important step in building underapproximations of string constraints. 
For two PFA $A$ and $A'$ over character variables $V$ and $V'$, respectively, 
it builds \emph{synchronisation formula} $\syncfof A {A'}$, a linear formula over $\cvarone\cup \cvartwo \cup \pvarone\cup\pvartwo$ encoding the semantic intersection $\semof A \cap \semof {A'}$. 
Besides the intersection itself, it also represents the set of all pairs of string encodings $(I\cup \pim)\in \parikhof {A}$ and $(I'\cup \pim')\in\parikhof {A'}$ such that 
$\decode A(I\cup \pim) = \decode {A'}(I'\cup \pim')$ which constitute the intersection.

Particularly, a synchronisation formula is build for two PFA  
$A = (Q,T,\cvarone,q_i,q_f)$ and ${A'} = (Q',T',\cvartwo,q_i',q_f')$ such that $\cvarone \cap \cvartwo = \emptyset$. 
%
It is extracted from the \emph{asynchronous product} of $A$ and $A'$. 
%
The asynchronous product is an automaton that uses $Q\times Q'$ as the set of states and 
$\cvarone_\epsilon\times\cvartwo_\epsilon$
as the alphabet. 
%
Every accepting run of $\syncof A {A'}$ corresponds to a pair of accepting runs, 
a run of $A$ over a parametric string $x$ and a run of ${A'}$ over a parametric string $x'$. 
The word accepted by the run of $\syncof A {A'}$ induces constraints on the interpretations of $I$ of $\cvarone$ and $I'$ of $\cvartwo$ under which the two parametric strings have the same interpretation, i.e. $I(x) = I(x')$.

@@

Formally, the asynchronous product automaton is a tuple $\syncof A {A'} = (Q\times Q', \syncT, \cvarone_\epsilon \times \cvartwo_\epsilon, (q_i,q_i'),(q_f,q_f'))$, where the transition relation $\syncT$ is the minimal set satisfying the following:
%
\begin{itemize}
\item If $(q_1,v,q_2) \in T$ and $(q_1',v',q_2') \in T'$, then we have $((q_1,q_1'),(v,v'),(q_2,q_2'))\in \syncT$.
\item If $(q_1,v,q_2) \in T$, then for all states $q'\in Q'$, we have $((q_1,q'),(v,\epsilon),(q_2,q'))\in \syncT$.
\item If $(q_1',v',q_2') \in T'$, then for all states $q\in Q$, we have $((q,q_1'),(\epsilon,v'),(q,q_2'))\in \syncT$.
\end{itemize}	
%
%The we define of constraints that extract from $\syncof A {A'}$ pairs of united assignments (encodings of strings over $\Sigma$) that encode the same string. 
The synchronisation formula $\syncfof{A} {A'}$ is extracted from $\syncof A {A'}$ as follows.
Its first part is the Parikh formula $\parikhfof {\syncof A {A'}}$ of the product, which encodes all runs of the product.
%
The second part is a constraint that extracts from a run of $\syncof A {A'}$ the corresponding runs of $A$ and of $A'$:
$$ \Psi_{\#} \defeq 
\left(\bigwedge_{v\in\cvarone}\#v = \sum_{x' \in \cvartwo_\epsilon} \#{(v,x')}\right)
\land
\left(\bigwedge_{v'\in\cvartwo} \#v' = \sum_{x \in \cvarone_\epsilon} \#{(x,v')}\right)
$$
Notice that $x,x'$ are either variables or $\epsilon$.
%\begin{enumerate}
%	\item for all $v \in \cvarone$, $\#(v) = \sum_{v' \in \cvartwo} \#{(v,v')}$   
%	\item for all $v \in \cvartwo$, $\#(v) = \sum_{v' \in \cvarone} \#{(v',v)}$ 
%\end{enumerate}
Finally, the third part forces the interpretations of the parametric strings accepted by $A$ and $A'$ be the same:
% condition 1. and 2. required for the pair of symbolic string $(w_n, w_m)$ and $I$, i.e., $I(w_n) = I(w_m)$ and $I$ assign the same value to all versions of the same variable (we use $-1$ as the value of $\epsilon$). 
$$ \Psi_= \defeq
\bigwedge_{x\in\cvarone_\epsilon,x'\in\cvartwo_\epsilon} \#(x,x')>0 \rightarrow (x=x')
$$
%\begin{enumerate}
%	\item for all $v\in \cvarone$ and $v'\in \cvartwo$, $\#{(v,v')}>0 \rightarrow (v=v')$, 
%	\item for all $v\in \cvarone$, $\#{(v,\epsilon)}>0 \rightarrow (v=\epsilon)$, and
%	\item for all $v\in \cvartwo$,$\#{(\epsilon,v')}>0 \rightarrow (v'=\epsilon)$.
%\end{enumerate}
The synchronisation formula is then the conjunction 
$$
%\syncfof A {A'} = \exists \pvarsof{\cvarone\times\cvartwo}\exists {(\cvarone\times\cvartwo)}.\,\parikhfof{\syncof A {A'}} \land \Psi_{\#} \land \Psi_=
\syncfof A {A'} \defeq \parikhfof{\syncof A {A'}} \land \Psi_{\#} \land \Psi_=
$$

The correctness of this construction is expressed by Lemma~\ref{lemma:synccorrect} below. The lemma can also be understood as a precise specification the semantics of the synchronisation formula, on which we rely for correctness of the construction of underapproximations of string constraints. 
%
%\newcommand\restrict[1]{\raisebox{-.5ex}{$\big|$}_{#1}}
\begin{lemma}\label{lemma:synccorrect}
%$\semof {\exists V_\times . \syncfof A {A'}}\restrict{\cvarone \cup \cvartwo \cup \pvarsof{\cvarone} \cup \pvarsof\cvartwo} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
$\semof {\exists V_\times . \syncfof A {A'}} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
\end{lemma}
%
Informally, the lemma states that the models of $\syncfof A {A'}$ encode precisely the pairs of equivalent encodings of words from $\semof{A}$ and $\semof{B}$, that constitute the intersection $\semof{A} \cap \semof{B}$. A minor technical difficulty is that the modes of $\syncfof A {A'}$ include also an assignment to the auxiliary variables of $V_\times = (\cvarone_\epsilon\times\cvartwo_\epsilon) \cup \pvarsof{(\cvarone_\epsilon\times\cvartwo_\epsilon)}$. Hence in the lemma, we have to restrict the models to the original variables $\cvarone \cup \cvartwo \cup \pvarsof{\cvarone} \cup \pvarsof\cvartwo$. This is  achieve by quantifying out the variables of $V_\times$.

>>>>

\begin{lemma}[\cite{SeidlSMH04}]
$\parikhof{A}$ of an FA $A$ can be computed in the form of a linear formula  $\parikhfof{A}$ such that $\modelsof{\parikhfof A}_{\#\Sigma} = \parikhof{A}$, in time linear to the size of $A$.
\end{lemma}

@@

%Such automata are recognizers of symbolic strings, i.e., strings over $\cvar$.
Parametric automata accept strings over $\cvar$, called \emph{parametric strings}, but we still use them as representations of languages over $\Sigma$. 
%
Namely, string over $\cvar$ are interpreted as strings over $\Sigma$ given an assignment of characters to its character variables.  
%
Formally, an \emph{interpretation $I$ over $\cvar$} is a mapping from $\cvar$ to $\Sigma_\epsilon$ and for a parametric string $x= v_1v_2\cdots v_k$ over $\cvar$, $I(x)$ is defined as $I(v_1)\cdot I(v_2)\cdot \ldots \cdot I(v_k)$.
%
We define the semantics of $x$ as the set $\semof x$ of all possible interpretations of $x$, the semantics of a language $L\subseteq V^*$ is the union $\semof L = \cup_{x\in L} \semof{x}$ of semantics of its words, and the semantics $\semof A$ of an PFA is the semantics of its language.

The crucial feature of parametric flat automata is that their semantics can be faithfully represented by a linear formula and handled efficiently by an SMT solver. 
\lh{, allowing to circumvent more costly automata algorithms and decision procedures. Concretize and justify this better?}\yfc{The two sentences below read a bit strange, maybe need to rephrase it.}
%
Achieving this was the purpose of introducing the structural constraints of PFA. 
Their implication central for our method is that every parametric string $x\in L(A)$ is uniquely determined by its Parikh image $\parikhof{x}$, as formalized by the following lemma.

@@

\yfc{Maybe we use things like $I_{\#}$ to replace $\pim$? It reads a bit strange when you put together $I$ and $\pim$, one is greek symbol and the other is roman, but both of them are interpretations.}

Since every string $w \in \semof{A}$ is obtained from a parametric string $x\in L(A)$ and an interpretation of character variables $I:\cvar\rightarrow\Sigma_\epsilon$ such that $w = I(x)$, and since $x = \decode A(\parikhof x)$, 
we may say that the assignment $I \cup \parikhof x$ uniting the interpretation $I$ and the Parikh image of the parametric string $x$ encodes $w$. Lemma~\ref{lemma:parikh} also implies that two different strings are encoded by different united assignments (the Parikh image or the interpretation must be different), and hence we may extend the decoding function $\decode A$ to united assignments so that $\decode A(I \cup \parikhof x) = w$.
Encodings of strings in $\semof{A}$ are encodings of parametric strings of $L(A)$ paired with arbitrary interpretations of character variables $\cvar$, which are precisely all models of the formula $\parikhfof A$ interpreted over $\cvar\cup\pvarsof \cvar$, as formulated by the following corollary of Lemma~\ref{lemma:decx}:
%
\begin{corollary}
For an PFA $A$ over character variables $V$,
$\semof{A} = \{\decode A(I\cup \pim) \mid I\cup\pim\in\semof{\parikhfof A}_{\cvar\cup\pvar}\}$.
\end{corollary}
%
A minor technical difficulty is that we wish to work with pure arithmetic formulas, 
but we may also wish to specify when a character variable is assigned $\epsilon$, which is not a number. 
Hence, to allow purely arithmetic formulas to talk about the value $\epsilon$, we encode $\epsilon$ as $-1$.
In the further development, $\epsilon$ will mean $-1$ when working with arithmetic formulas constraining interpretations of character variables.

\paragraph{Synchronization of PFA}
The following construction is an important step in building under-approximations of string constraints. 
For two PFA $A$ and $A'$ over character variables $V$ and $V'$, respectively, 
it builds \emph{synchronization formula} $\syncfof A {A'}$, a linear formula over $\cvarone\cup \cvartwo \cup \pvarone\cup\pvartwo$ encoding the semantic intersection $\semof A \cap \semof {A'}$. 
Besides the intersection itself, it also represents the set of all pairs of string encodings $(I\cup \pim)\in \parikhof {A}$ and $(I'\cup \pim')\in\parikhof {A'}$ such that 
$\decode A(I\cup \pim) = \decode {A'}(I'\cup \pim')$ which constitute the intersection.

Particularly, a synchronization formula is build for two PFA  
$A = (Q,T,\cvarone,q_i,q_f)$ and ${A'} = (Q',T',\cvartwo,q_i',q_f')$ such that $\cvarone \cap \cvartwo = \emptyset$. 
%
It is extracted from the \emph{asynchronous product} of $A$ and $A'$. 
%
The asynchronous product is an automaton that uses $Q\times Q'$ as the set of states and 
$\cvarone_\epsilon\times\cvartwo_\epsilon$
as the alphabet. 
%
Every accepting run of $\syncof A {A'}$ corresponds to a pair of accepting runs, 
a run of $A$ over a parametric string $x$ and a run of ${A'}$ over a parametric string $x'$. 
The word accepted by the run of $\syncof A {A'}$ induces constraints on the interpretations of $I$ of $\cvarone$ and $I'$ of $\cvartwo$ under which the two parametric strings have the same interpretation, i.e. $I(x) = I(x')$.

@@

Formally, the asynchronous product automaton is a tuple $\syncof A {A'} = (Q\times Q', \syncT, \cvarone_\epsilon \times \cvartwo_\epsilon, (q_i,q_i'),(q_f,q_f'))$, where the transition relation $\syncT$ is the minimal set satisfying the following:
%
\begin{itemize}
\item If $(q_1,v,q_2) \in T$ and $(q_1',v',q_2') \in T'$, then we have $((q_1,q_1'),(v,v'),(q_2,q_2'))\in \syncT$.
\item If $(q_1,v,q_2) \in T$, then for all states $q'\in Q'$, we have $((q_1,q'),(v,\epsilon),(q_2,q'))\in \syncT$.
\item If $(q_1',v',q_2') \in T'$, then for all states $q\in Q$, we have $((q,q_1'),(\epsilon,v'),(q,q_2'))\in \syncT$.
\end{itemize}	
%
%The we define of constraints that extract from $\syncof A {A'}$ pairs of united assignments (encodings of strings over $\Sigma$) that encode the same string. 
The synchronization formula $\syncfof{A} {A'}$ is extracted from $\syncof A {A'}$ as follows.
Its first part is the Parikh formula $\parikhfof {\syncof A {A'}}$ of the product, which encodes all runs of the product.
%
The second part is a constraint that extracts from a run of $\syncof A {A'}$ the corresponding runs of $A$ and of $A'$:
$$ \Psi_{\#} \defeq 
\left(\bigwedge_{v\in\cvarone}\#v = \sum_{x' \in \cvartwo_\epsilon} \#{(v,x')}\right)
\land
\left(\bigwedge_{v'\in\cvartwo} \#v' = \sum_{x \in \cvarone_\epsilon} \#{(x,v')}\right)
$$
Notice that $x,x'$ are either variables or $\epsilon$.
%\begin{enumerate}
%	\item for all $v \in \cvarone$, $\#(v) = \sum_{v' \in \cvartwo} \#{(v,v')}$   
%	\item for all $v \in \cvartwo$, $\#(v) = \sum_{v' \in \cvarone} \#{(v',v)}$ 
%\end{enumerate}
Finally, the third part forces the interpretations of the parametric strings accepted by $A$ and $A'$ be the same:
% condition 1. and 2. required for the pair of symbolic string $(w_n, w_m)$ and $I$, i.e., $I(w_n) = I(w_m)$ and $I$ assign the same value to all versions of the same variable (we use $-1$ as the value of $\epsilon$). 
$$ \Psi_= \defeq
\bigwedge_{x\in\cvarone_\epsilon,x'\in\cvartwo_\epsilon} \#(x,x')>0 \rightarrow (x=x')
$$
%\begin{enumerate}
%	\item for all $v\in \cvarone$ and $v'\in \cvartwo$, $\#{(v,v')}>0 \rightarrow (v=v')$, 
%	\item for all $v\in \cvarone$, $\#{(v,\epsilon)}>0 \rightarrow (v=\epsilon)$, and
%	\item for all $v\in \cvartwo$,$\#{(\epsilon,v')}>0 \rightarrow (v'=\epsilon)$.
%\end{enumerate}
The synchronization formula is then the conjunction 
$$
%\syncfof A {A'} = \exists \pvarsof{\cvarone\times\cvartwo}\exists {(\cvarone\times\cvartwo)}.\,\parikhfof{\syncof A {A'}} \land \Psi_{\#} \land \Psi_=
\syncfof A {A'} \defeq \parikhfof{\syncof A {A'}} \land \Psi_{\#} \land \Psi_=
$$

The correctness of this construction is expressed by Lemma~\ref{lemma:synccorrect} below. The lemma can also be understood as a precise specification the semantics of the synchronization formula, on which we rely for correctness of the construction of underapproximations of string constraints. 
%
%\newcommand\restrict[1]{\raisebox{-.5ex}{$\big|$}_{#1}}
\begin{lemma}\label{lemma:synccorrect}
%$\semof {\exists V_\times . \syncfof A {A'}}\restrict{\cvarone \cup \cvartwo \cup \pvarsof{\cvarone} \cup \pvarsof\cvartwo} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
$\semof {\exists V_\times . \syncfof A {A'}} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
\end{lemma}
%
Informally, the lemma states that the models of $\syncfof A {A'}$ encode precisely the pairs of equivalent encodings of words from $\semof{A}$ and $\semof{B}$, that constitute the intersection $\semof{A} \cap \semof{B}$. A minor technical difficulty is that the modes of $\syncfof A {A'}$ include also an assignment to the auxiliary variables of $V_\times = (\cvarone_\epsilon\times\cvartwo_\epsilon) \cup \pvarsof{(\cvarone_\epsilon\times\cvartwo_\epsilon)}$. Hence in the lemma, we have to restrict the models to the original variables $\cvarone \cup \cvartwo \cup \pvarsof{\cvarone} \cup \pvarsof\cvartwo$. This is  achieve by quantifying out the variables of $V_\times$.
